---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

1. Load the data (i.e. read.csv())
```{r echo=TRUE}
unzip("activity.zip")
activity_data <- read.csv("activity.csv", header=TRUE, na.strings= "NA")
head(activity_data)
tail(activity_data)
```

2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r echo=TRUE}
class(activity_data$date) # Checking class of date variable in dataset

activity_data$date <- as.Date(activity_data$date) # converting date value to factor to date format

class(activity_data$date) # Checking class of date variable in dataset
```

## What is mean total number of steps taken per day?

1. Make a histogram of the total number of steps taken each day
```{r}
total_steps <- tapply(activity_data$steps, activity_data$date, FUN=sum) # Aggregating number of steps for each day

# Plotting histogram of total number of steps taken each day
hist(total_steps,
     breaks=11,
     freq=TRUE,
     main="Histogram of total number of steps taken each day",
     xlab="Total number of steps taken per day")
```

2. Calculate and report the mean and median total number of steps taken per day
```{r echo=TRUE}
mean_steps <- mean(total_steps, na.rm=TRUE)
median_steps<- median(total_steps, na.rm=TRUE)
```
```{r echo=TRUE}
mean_steps
median_steps
```


## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r echo=TRUE, message=FALSE, warning=FALSE}
avg_steps <- aggregate(x=list(steps=activity_data$steps),
              by=list(interval=activity_data$interval),
              FUN=mean,
              na.rm=TRUE)

plot(avg_steps$steps~avg_steps$interval,
     type="l",
     col="blue",
     main="5−minute interval versus average number of steps taken",
     xlab="5-minute interval",
     ylab="the average number of steps taken")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE}
avg_steps[which.max(avg_steps$steps),]
```
The maximum number of steps is 206.1698. The corresponding 5-minute interval identifier is 835. 


## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r echo=TRUE}
table(is.na(activity_data))
```
There are 2304 NA values in dataset.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Here, we replace the missing values by the mean for the corresponding 5-minute interval.
```{r echo=TRUE}
value_filled <- function(steps, interval) {
  filled <- NA
  if (!is.na(steps))
    filled <- steps
  else
    filled <- avg_steps[avg_steps$interval==interval, "steps"]
  return(filled)
}
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r echo=TRUE}
activity_data_new <- activity_data # Creating new dataset 
activity_data_new$steps <- mapply(value_filled, activity_data_new$steps, activity_data_new$interval)
# show that the missing values are gone
table(is.na(activity_data_new))
```

4. Histogram of the total number of steps taken each day using the new data set

```{r echo=TRUE}
total_steps_new <- tapply(activity_data_new$steps, activity_data_new$date, FUN=sum)

hist(total_steps_new,
     breaks=11,
     freq=TRUE,
     main="Histogram of total number of steps taken each day",
     xlab="total number of steps taken per day")
```

Mean and Median total number of steps taken per day
```{r echo=TRUE}
mean_new <- mean(total_steps_new, na.rm=TRUE)
median_new <- median(total_steps_new, na.rm=TRUE)
```

Let’s compare the new mean and median to those we calculated above
```{r echo=TRUE}
mean_steps <= mean_new
median_steps < median_new
```


## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r echo=TRUE}
activity_data_new$date <- as.Date(activity_data_new$date)
activity_data_new$weekdays <- factor(format(activity_data_new$date, "%A"))
levels(activity_data_new$weekdays) <- list(weekday = c("Monday", "Tuesday","Wednesday",
                                                       "Thursday", "Friday"),
                                           weekend = c("Saturday", "Sunday"))
```

Let's check new data
```{r echo=TRUE}
head(activity_data_new)
tail(activity_data_new)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r echo=TRUE}
library(lattice)
avg_steps_new <- aggregate(steps ~ interval + weekdays, data=activity_data_new, mean)
xyplot(avg_steps_new$steps ~ avg_steps_new$interval | avg_steps_new$weekdays,
       layout = c(1, 2),
       type = "l",
       xlab = "5-Minute Intervals",
       ylab = "Averaged Number of steps")

```
